Annotation exploration for SCPCP000006

### 2024-09-25 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
changes to PR#776

---
title: "Annotation exploration for `r params$scpca_project_id`"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

```{r packages, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)

@@

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", params$scpca_project_id)

@@

In this notebook, we are working with all of the samples in `r params$scpca_project_id`.

@@

```{r path_to_query}
sample_metadata_file <- file.path(repository_base, "data", "current", params$scpca_project_id, "single_cell_metadata.tsv")

@@

#### do_Table_Heatmap

`do_Table_Heatmap` shows heatmap of counts of cells for combinations of two metadata variables

- `first_group` is the name of the first metadata to group the cells
- `last_group` is the name of the second metadata to group the cells

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Table_Heatmap <- function(first_group, last_group ){

  df <- cell_type_df %>%
  group_by( !!sym(first_group), !!sym(last_group))%>% 
  summarise(Nb = n())  %>%
  mutate(C = sum(Nb)) %>%
  mutate(percent = Nb/C*100)

p1 <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = Nb)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   guides(fill=guide_legend(title=" Number of \n cells"))

p2 <- ggplot(df, aes(x= !!sym(first_group), y = !!sym(last_group), fill = percent)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   guides(fill=guide_legend(title=" Percentage of \n cells"))

p <- p1 + p2 + plot_layout(ncol = 2)
return(p)
}
```

@@

- `group.by` is the name of the metadata to group the cells
- `feature` is the name of the gene to average the expression

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_mean <- function(group.by, feature ){

  df <- cell_type_df %>%
  group_by(sample_id, !!sym(group.by))%>% 
  summarise(m = mean(!!sym(feature)))

p <- ggplot(df, aes(x= sample_id, y = !!sym(group.by), fill = m)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   guides(fill=guide_legend(title=paste0(feature)))

@@

### Total number of cells per samples

```{r fig.width=5, fig.height=10, out.width='100%', results='asis'}
tmp <- table(cell_type_df$sample_id)
knitr::kable(tmp)
```
### Predicted organ

@@

```{r fig.width=25, fig.height=6,  results='asis'}
tmp <- table(cell_type_df$sample_id, cell_type_df$organ)

knitr::kable(tmp)
```

```{r fig.width=25, fig.height=6}
do_Table_Heatmap(first_group = "sample_id", last_group = "organ")
```

@@

```{r fig.width=25, fig.height=3, results='asis'}
tmp <- table(cell_type_df$sample_id, cell_type_df$compartment)

knitr::kable(tmp)
```

```{r fig.width=25, fig.height=3}

do_Table_Heatmap(first_group = "sample_id", last_group = "compartment")
```

@@

```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(  feature = "ENSG00000110799",
                  group.by = "compartment")
```

@@

```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(  feature = "ENSG00000081237",
                  group.by = "compartment")
```

>>>>

---
title: "Annotation exploration for SCPCP000006"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

```{r packages, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(tidyr)
library(patchwork)
library(DT)
```

@@

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", "SCPCP000006")

@@

In this notebook, we are working with all of the samples in SCPCP000006.

@@

```{r path_to_query}
sample_metadata_file <- file.path(repository_base, "data", "current","SCPCP000006", "single_cell_metadata.tsv")

@@

@@

- `df` is the name of the table containing metadata and feature expression (counts) per cells
- `group.by` is the name of the metadata to group the cells
- `feature` is the name of the gene to average the expression

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_mean <- function(df, group.by, feature ){

  df <- df %>%
  group_by(sample_id, !!sym(group.by))%>% 
  summarise(m = mean(!!sym(feature)))

p <- ggplot(df, aes(x= sample_id, y = !!sym(group.by), fill = m)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) +
   guides(fill=guide_legend(title=paste0(feature)))

#### do_Feature_boxplot

`do_Feature_boxplot` shows boxplot of expression of a feature grouped by a metadata.

- `table` is the name of the table containing metadata and feature expression (counts) per cells
- `group.by` is the name of the metadata to group the cells
- `feature` is the name of the gene to average the expression

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_boxplot <- function(df, group.by, feature, split.by ){

  df <- df %>%
    mutate(!!sym(group.by), group = factor(!!sym(group.by), levels = unique(!!sym(group.by))))

  xlabs <- paste(levels(df$group),"\n(N=",table(df$group),")",sep="")

p <- ggplot(df, 
    aes(x = !!sym(group.by), y = !!sym(feature), fill = !!sym(group.by))
  ) + 
  geom_boxplot(size = 0.5, size.outlier = 0.25) +
  facet_wrap(vars(!!sym(split.by)), scales = "free", ncol = 4) + 
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12, angle = 0, hjust = 0.5, vjust = 0), 
    legend.position = "none"
  ) +scale_x_discrete(labels=xlabs)

@@

### Total number of cells per samples and predicted organ

@@

Here, we are summarizing per sample:

- the percentage of cells labeled as kidney, after label transfer from the fetal full reference [Cao et al.](https://www.science.org/doi/10.1126/science.aba7721),

- the percentage of cells not labeled as kidney, after label transfer from the fetal full reference [Cao et al.](https://www.science.org/doi/10.1126/science.aba7721),

- the total number of cells per sample.

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
kidney_count_df <- cell_type_df |>
  # make a new variable that tells us if it's kidney or not
  dplyr::mutate(organ_type = ifelse(organ == "Kidney", "kidney", "not_kidney"))  |>
  # count how many kidney or not
  dplyr::count(sample_id, organ_type) |>
  # make the data frame wide and use a value of 0 if a count is missing
  tidyr::pivot_wider(names_from = organ_type, values_from = n, values_fill = 0) |>
  # add a column for total, and convert other counts to percentages
  dplyr::mutate(
    total = kidney + not_kidney, 
    kidney = round(kidney/total, digits = 4)*100, 
    not_kidney = round(not_kidney/total, digits = 4)*100) |>
  # arrange in descending order of percentage of kidney cells
  dplyr::arrange(desc(kidney))

DT::datatable(kidney_count_df, 
              caption = "percentage of cells mapping to the predicted organ kidney",
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
                               buttons = c( 'csv', 'excel')))
```

@@

```{r fig.width=25, fig.height=3, results='asis'}
compartment_df <- cell_type_df |>
  # count how many per compartment
  dplyr::count(sample_id, compartment) |>
  # make the table wide and use a value of 0 if a count is missing
  tidyr::pivot_wider(names_from = compartment, values_from = n, values_fill = 0) |>
  # order the columns how we want to show them
  dplyr::select(sample_id, fetal_nephron, stroma, endothelium, immune)

DT::datatable(compartment_df, 
              caption = "counts of cells in each compartment",
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
                               buttons = c( 'csv', 'excel')))
```

@@

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df,
                  feature = "ENSG00000110799",
                  group.by = "compartment")
```

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_boxplot( df = cell_type_df,
                    feature = "ENSG00000110799",
                    group.by = "compartment",
                    split.by = "sample_id")

@@

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(  df = cell_type_df,
                  feature = "ENSG00000081237",
                  group.by = "compartment")
```

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_boxplot( df = cell_type_df,
                    feature = "ENSG00000081237",
                    group.by = "compartment",
                    split.by = "sample_id")
